Scheduling and Parallelization for MPC

### 2019-11-20 paper.tex
Just checking if this works.
some edits

We now consider the answer to Problem 2 (in MPC).

\subsection{Problem Overview}

Recall Problem 2, given a loop-body schedule, compute the across-loop schedule
with minimal cost. The following examples illustrate. The first example is the inner 
product computation, and the second is Aiken and Nicolau's example.

@@

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=1; i<=N; i++) {
  if (i%step5==1) para(t[i],...,t[i+step])
  if (i%step6==1) ip = ip + t[i];
}  
\end{lstlisting}

@@

In the above example {\sf step5} is N, meaning that all N multiplications are executed in parallel.
In contrast, {\sf step6} is 1, meaning that the summation statements are executed sequentially.

@@

The above across-loop schedule means that the first instruction is parallelized---all
N instructions A can be computed at once because each depends only on {\sf B[i]}, which
is assigned outside of the loop. In contrast, subsequent instructions are executed 
sequentially. Take B, for example. There is a dependence ``cycle'' from {\sf B[i]} 
to {\sf D[i]} then {\sf B[i+1]} in the subsequent iteration. (The cycle is shown in~\figref{fig:MPCexample}(b).) 
{\sf B[i]}'s cannot be computed in parallel because each one depends on the {\sf B[i]} 
value computed in the previous iteration.

Now consider the above code but slightly modified. {\sf B[i]} and {\sf C[i]} are now 
computed based on the value of {\sf D[i]} from two iterations ago.

@@

{\sf B[i]} still depends on values computed in earlier iterations, but one of the values, D[i-2] 
is two iterations earlier. Thus, we can take parallelization steps of 2, e.g., computing 
{\sf B[1]} and {\sf B[2]} in parallel, then {\sf B[3]} and {\sf B[4]} in parallel, etc.
Note however, that we cannot take a parallelization step larger than 2.
The value of B will depend on a value being computed in parallel.

To simplify, we assume that a loop-body schedule is a straight line. \ana{What does this mean?}
\ana{Will we relax?}
The problem therefore becomes: compute the maximal parallelization 
step allowed for each statement. This will account for an MPC schedule with
minimal cost. \ana{Why?}

@@

\ana{1. MPC-Source}

@@

(a) Source 
&
(c) MPC-source

@@

\paragraph{MPC-Source.} MPC-Source is an SSA-based intermediate representation
of MPC programs~\cite{ccs paper}. MPC-Source is, essentially, a straight-line sequence
of ``components'', where each ``component'' is either a nested for-loop sequence (called for-loop block), 
or a simple MPC-statement. \figref{fig:example} shows an example. MPC-Source can be constructed
in a standard way from an SSA-translation of the source---each branch of an 
{\sf if}-statement is added to the sequence, the conditional becomes an MPC comparison (CMP), 
and the $\phi$-node becomes an MPC multiplexer (MUX).

\paragraph{Array SSA.} The other essential concept we build upon is Array SSA. Originally proposed

@@

\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

@@

  A2[i] = f1(B1[i]); @A2[i] = i;
  A3 = `$\phi$'(A2,A1); @A3 = max(A2, A1);
  B2[i] = f2(A3[i],D1[i-1]); @B2[i] = i;
  B3 = `$\phi$'(B2,B1); @B3 = max(B2, B1);
  C3[i] = f3(A3[i],D1[i-1]); @C2[i] = i;
  C3 = `$\phi$'(C2,C1); @C3 = max(C2, C1);
  D3[i] = f4(B3[i],C3[i]); @D2[i] = i;
  D3 = `$\phi$'(D2,D1); @D3 = max(D2, D1); 
}
\end{lstlisting}

"Index invariant" makes an assertion about when A[i] is written as a function of i.
\ana{Figure out how to reason about cycles based on Array SSA.}

\subsection{Algorithm}

@@

>>>>

We now consider the answer to Problem 2 (in MPC). 
\ana{Make a note that the problem is easy assuming NO array Writes. We consider array writes.}

\subsection{Overview and Problem Statement}

The problem is, given a loop-body schedule, compute the across-loop schedule
with minimal cost. The following examples illustrate. The first example is the inner 
product computation, and the second is Aiken and Nicolau's example.

@@

\end{lstlisting}
& 
~
&
\begin{lstlisting}[language=Java]
A = ...
B = ...
ip = 0;
for (i=1; i<=N; i++) {
  if (i%d(5)==1) para(t[i],...,t[i+d(5)])
  if (i%d(6)==1) ip = ip + t[i];
}  
\end{lstlisting}

@@

In the above example {\sf d(5)} is N, meaning that all N multiplications are executed in parallel.
In contrast, {\sf d(6)} is 1, meaning that the summation statements are executed sequentially.

@@

The above across-loop schedule entails that the first instruction is parallelized---all
N instructions A can be computed at once because each depends only on {\sf B[i]}, 
which at the point {\sf A[i]} is computed, refers to a value assigned outside of the loop. 
In contrast, subsequent instructions are executed sequentially. Take B, for example. 
There is a dependence ``cycle'' from {\sf B[i]} to {\sf D[i]} then {\sf B[i+1]} in the subsequent 
iteration. (The cycle is shown in~\figref{fig:MPCexample}(b).) {\sf B[i]}'s cannot be computed 
in parallel because each one depends on the {\sf B[i]} value computed in the previous iteration.

Now consider the above code but slightly modified. {\sf B[i]} and {\sf C[i]} are now 
computed based on the value of {\sf D[i-2]} assigned two iterations earlier.

@@

{\sf B[i]} still depends on values computed in earlier iterations, but one of the values, D[i-2] 
is two iterations earlier. Thus, we can take parallelization steps of 2, e.g., computing 
{\sf B[1],B[2]} in parallel, then {\sf B[3],B[4]}, etc.
Note however, that we cannot take a parallelization step larger than 2.
%The value of B will depend on a value being computed in parallel.

To simplify, we assume that a loop-body schedule is a straight line. \ana{What does this mean?}
\ana{Will we relax?}
The problem becomes: for each step, compute the maximal parallelization 
step allowed by program dependences. This will account for an MPC schedule with 
minimal cost. \ana{Why?}

@@

@@

(a) GCD Source 
&
(c) MPC-source

@@

\subsubsection{MPC-Source}

MPC-Source is an SSA-based intermediate representation
of MPC programs~\cite{ccs paper}. MPC-Source is, essentially, a straight-line sequence
of ``components'', where each ``component'' is either a nested for-loop sequence (called for-loop block), 
or a simple MPC-statement. \figref{fig:example} shows an example. MPC-Source can be constructed
in a standard way from an SSA-translation of the source---each branch of an 
{\sf if}-statement is added to the sequence, the conditional becomes an MPC comparison (CMP), 
and the $\phi$-node becomes an MPC multiplexer (MUX).

\ana{Add a paragraph on handling of arrays in standard SSA.}

\subsubsection{Array SSA}

The other essential concept we build upon is Array SSA. 
Originally proposed by Knobe and Sarkar, Array SSA has been used for loop parallelization in HPC.
We adapt Array SSA for MPC-source and build our optimal loop parallelization algorithm
based on the adaptation.

@@

Array SSA is best illustrated by an example. Consider Aiken and Nicolau's example in Array SSA:

\begin{figure}
\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

@@

  A2[i] = f1(B1[i]); @A2[i] = i;
  A3 = `$\phi$'(A2,A1); @A3 = max(A2, A1);
  B2[i] = f2(A3[i],D1[i-1]); @B2[i] = i;
  B3 = `$\phi$'(B2,B1); @B3 = max(B2, B1);
  C3[i] = f3(A3[i],D1[i-1]); @C2[i] = i;
  C3 = `$\phi$'(C2,C1); @C3 = max(C2, C1);
  D3[i] = f4(B3[i],C3[i]); @D2[i] = i;
  D3 = `$\phi$'(D2,D1); @D3 = max(D2, D1); 
}
\end{lstlisting}
\caption{Aiken and Nicolau's example using Array SSA.}
\label{fig:ArraySSAAikenNicolau}
\end{figure}

There are three kinds of statements added by Array SSA. First, 
\[
{\sf A2[i] = f1(B1[i]); @A2[i] = i;}
\]
Array SSA statement {\sf @A2[i] = i;} records the iteration, {\sf i} when {\sf A2[i]} is written.
Note that the index is not necessarily {\sf i}, it can be any $f(i)$, however, we assume
that the loop always iterates from 1 to $N$ with step 1. Second and third,
\[ 
{\sf A3 = }\phi{\sf (A2,A1); @A3 = max(A2, A1);}
\]
Statement ${\sf A3 = }\phi{\sf (A2,A1);}$ is a \emph{define} $\phi$ node where arrays 
{\sf A2} and {\sf A1} are ``merged''---in this case, each value at index other than $i$ 
is taken from array {\sf A1}, and index $i$ is taken from {\sf A2}. Statement 
{\sf @A3 = max(A2, A1);} is a timestamp statement; it picks the latest iteration 
to assign for each index $i$---in this case, when the phi- and max- statements 
appear immediately under an array write, it assigns {\sf @A3[j] = i} if $j=i$, 
and {\sf @A3[j] = @A2[j]} otherwise.

\ana{Add merge $\phi$}

\subsubsection{MPC-Source with Array SSA}

The phi-node at if-then-else statements, e.g., ${\sf A3 = \phi(A2,A1), @A3 = max(A2, A1);}$ 
works for MPC-source without loss of privacy. ${\sf A3 = \phi(A2,A1)}$ is translated as 
point-wise MUX for each index $i$. The timestamp statement ${\sf @A3 = max(A2, A1);}$ 
has the same meaning as in standard Array SSA---{\sf @A3[i] = max(@A2[i],@A1[i])}.
Since MPC-source expands and executes each arm of the if-then-else statement,
if the then-arm modifies the array at index $i$, but the else arm does not, 
{\sf @A3[i] = max(@A2[i],@A1[i]) = i}. Similarly, if both arms modify the array at $i$,
the index again will be $i$. We note that in MPC-source the timestamp arrays {\sf @A} 
are known \emph{statically}, unlike in standard Array SSA, where they are inherently \emph{dynamic}. 
\ana{Double check? Make a statement that an adversary monitoring execution does not
infer anything from SSA phi nodes.}

\subsection{Analysis Algorithm}

\subsubsection{Restrictions on MPC-Source}

To make the problem tractable we adopt certain constraints on the loop body and loop body schedule.
These constraints hold in classical works on optimal loop parallelization, e.g., Aiken and Nicolau.

Array writes.~\ana{We have to see what the distinction was, regular vs. irregular loops}:
\begin{enumerate}
\item For each array read of $A$: $... = ... A[f(i)]$ and corresponding write to $A$: $A[f'(i)] = ...$, 
we have that $f(i) = f'(i)$ for every $0 \le i < N$. Here $i$ is the induction variable and $N$ is
the loop bound.

\item Let $A[f_1(i)]...A[f_k(i)]$ be a sequence of writes to $A$ in the loop body sequence 
\ana{How is the sequence determined?} with $A[f_k(i)]$ being the most recent one. Here
$i$ denotes the induction variable. We require that for every $0 \le i < N, j,k$, $f_j(i) \neq f_k(i)$, 
or in other words, those writes access different elements of the array. Furthermore, we require 
that $f_k(i) \neq f_l(j)$, or in other words, writes across different iterations access different array 
elements. \ana{Needs rewording. Can be simplified.}
\end{enumerate}

Constant loop schedule. Let $S$ be a loop schedule, i.e., a directed acyclic graph that defines "executes before"
scheduling constraints among the statements of the loop body. We require that the 
parallelization of the loop schedule obeys the constant loop schedule constraint: each instance of $S$
in the loop linearization obeys the scheduling constraints defined by the DAG of $S$. \ana{Justify why!}

\subsubsection{Computing the Index Invariant}

"Index invariant" makes an assertion about when A[i] is written as a function of i.
\ana{Can we compute this index easily: A[f(j)] = i, if j>=i, A[f(j)] = 0 otherwise?
Under what assumptions this is true???}. \ana{We will use the Array SSA stuff to compute index; 
can get rid of it once we.}

\subsubsection{Computing $d(s)$}
We assume a loop body that consists of three-address MPC statements:
\[ v1 = v2 \oplus v3 \]
MUX statements
\[ v1 = MUX(v2, v3, v4) \]
and
phi-nodes
\[ v1 = \phi(v2, v3) \]
where $v1,v2$, $v3$, and $v4$ are either scalar variables, e.g., {\sf x} or array elements {\sf A[i]}.
Array-write phi- and max- statements are elided from the representation, as
their only purpose is to compute the invariant. Control-merge phi- are translated
into point-wise MUX statements, as is standard in MPC. \ana{Are they translated during analysis?}

\ana{Pseudo phi nodes!}

\begin{figure}
\begin{lstlisting}[language=Java,escapeinside=`']
for (i=1; i<=N; i++) {

  // Index Invariant: @A1[j] == j if `$j < i$', @A1[j] == 0 otherwise

  A1 = `$\phi$'(A3,A0); 
  B1 = `$\phi$'(B3,B0); 
  C1 = `$\phi$'(C3,C0);
  D1 = `$\phi$'(D3,D0);

  A2[i] = f1(B1[i]); 
  B2[i] = f2(A3[i],D1[i-1]); 
  C3[i] = f3(A3[i],D1[i-1]); 
  D3[i] = f4(B3[i],C3[i]); 
}
\end{lstlisting}
\caption{Aiken and Nicolau's example using Array SSA.}
\label{fig:ArraySSAMPCAikenNicolau}
\end{figure}

%\ana{Figure out how to reason about cycles based on Array SSA.}

Recall that the loop body is a linear sequence of statement. To compute $d(s)$ of statement $s$, 
we follow def-use chains backwards collecting all reachable pseudo-phi node. If there is a psudo-phi 
node of a scalar, then $d(s) = 1$; for an array we retrieve the value $A[f(i)]$ from the invariant. \ana{formalize, polish!}

\ana{Define Schedule $S$.}

\ana{Define "constant loop body schedule"}

\ana{Prove main theorem, that this schedule is of minimal cost. I.e., under the restriction of "constant loop body schedule", there
does not exist a schedule $S'$}

@@

\section{Extensions}

\subsection{Multi Dimensional Arrays}

\subsection{Nested Loops}
